Highly Available Transactions: Virtues and Limitations

### 2013-04-01 writeup/conclusion.tex
editing pass

\section{Conclusion}
The current state of database software offers uncomfortable and
unnecessary choices between availability and transactional semantics.
We show, however, through analysis and experiments, that the
availability goals of the NoSQL movement will remain a critical aspect
of many important systems going forward. Thus, we expose a broad
design space of Highly Available Transactional Systems (HATs), which
can offer the key benefits of highly available distributed
systems---partition tolerance and low-latency writes---while also
providing a family of transactional isolation levels and replica
semantics that have been widely used in practice.  In taxonomizing
this design space, we show that many transactional guarantees are
achievable with high availability, while a few, like preventing Lost
Update and Write Skew, are not. From a research perspective, this
offers a long-overdue connection between storage semantics studied in
the database and distributed systems communities.

With the above understanding in hand, the next challenge is to return to
issues of performance and systems architecture within the HATs design
space.  Based on our understanding of what is desired in the field and
what is possible to achieve, we believe this is an area of great
opportunity in the coming years.

>>>>

\section{Conclusion}
The current state of database software offers uncomfortable and
unnecessary choices between availability and transactional semantics.
We show, however, through analysis and experiments, that the
availability goals of the NoSQL movement will remain a critical aspect
of many future data storage systems. Accordingly, we expose a broad
design space of Highly Available Transactional Systems (HATs), which
can offer the key benefits of highly available distributed
systems---''always on'' operation during partitions and low-latency
operations---while also providing a family of transactional isolation
levels and replica semantics that have been adopted in practice.  In
taxonomizing this design space, we show that many transactional
guarantees are achievable with high availability, while a few, like
preventing Lost Update and Write Skew, are not. From a research
perspective, this offers a long-overdue connection between storage
semantics studied in the database and distributed systems communities.

With the above understanding in hand, the next challenge is to return
to issues of performance and systems architecture within the HAT
design space.  Based on our understanding of what is desired in the
field and newfound knowledge of what is possible to achieve, we
believe this is an area of great opportunity in the coming years.

### 2013-04-01 writeup/hat-analysis.tex
editing pass

``Repeatable Read'' isolation is a confusing property. Some IBM
products use the term for fully serializable
isolation~\cite{hat-hotos}. Gray~\cite{gray-isolation}, Berenson et
al.~\cite{ansicritique}, and Adya's PL-2.99~\cite{adya} all interpret
Repeatable Read as providing serializability for all operations except
predicate-based reads (and so Repeatable Read is identical to
serializability in a key-value store without predicate-based
access). In this model, the \textit{Phantom Problem}, whereby two
successive predicate-based reads return different
data~\cite{gray-isolation}, is the only kind of non-serializable
behavior allowed. When referring to ``Repeatable Read,'' we will
follow these definitions, which, as we will soon show, are for a
property that be offered in a highly available way.

@@

A stronger achievable property asks that a consistent cut spans both a
transaction's predicate reads (e.g., \texttt{SELECT WHERE}) as well as
its item reads.  We call this \textit{Predicate Cut
  Isolation}\footnote{Oracle provides an isolation level called
  ``Statement Level Read Consistency''~\cite{adya}; this is analogous
  to Predicate Cut Isolation at the level of a single operation within
  a multi-operation transaction. We do not consider this isolation
  model here as it is subsumed by standard Snapshot Isolation and we
  believe our discussion is easily extended to incorporate it.} and it
prohibits Phantoms. Predict Cut Isolation is also achievable in HAT
system via similar multi-versioning as above.

@@

Perplexingly, discussions of TA are absent from existing discussions
of weak isolation. This is perhaps again due to the single-node
context in which prior work was developed: on a single server (or a
fully replicated database), TA is achievable via lightweight locking
and/or local concurrency control over data items~\cite{gstore}. In
contrast, in a distributed environment, TA over arbitrary groups of
non-colocated items is considerably more difficult to achieve with
high availability: servers need to know when all of a transaction's
updates are present on their respective replicas. However, replicas do
not need to agree on when to reveal a new value to clients, which
would require consensus; accordingly, TA only requires reliable
broadcast (with some additional metadata for clients) and is
achievable in a HAT system. We omit a full discussion of the algorithm
and instead refer the reader to the extended version of this
paper~\cite{hat-tr}.  space{.5em}

@@

In this section, we discuss several subleties in our results,
specifically addressing model composition, transactional atomicity
versus linearizability, and stickiness requirements.

@@

\vspace{.5em}\noindent\textbf{Linearizability and Transactional
  Atomicity} The relationship between linearizability and
transactional atomicity is non-obvious. Transactional atomicity
dictates that writes to multiple keys across multiple servers are made
visible to readers all at once, while linearizability dictates that
writes to a single key on multiple servers are made visible to all
readers at once---what is different? First, in linearizable (and safe
and regular) systems, writes are made visible to all clients
\textit{immediately} after they finish. With transactional atomicity,
there is no recency guarantee. Second, in linearizabile systems, all
clients see all writes at the same time. With transactional atomicity,
clients may see writes at different times depending on which replicas
they contact. We are not aware of an analogous model in the
distributed systems literature. Accordingly, despite apparent
similarities, transactional atomicity incomparable with and much
cheaper (by availability standards) than linearizability.

>>>>

``Repeatable Read'' isolation is a confusing property. Some IBM
products use the term for fully serializable
isolation~\cite{hat-hotos}. Gray~\cite{gray-isolation}, Berenson et
al.~\cite{ansicritique}, and Adya's PL-2.99~\cite{adya} all interpret
Repeatable Read as providing serializability for all operations except
predicate-based reads (and so Repeatable Read is identical to
serializability in a key-value store without predicate-based
access). In this model, the \textit{Phantom Problem}, whereby two
successive predicate-based reads return different
data~\cite{gray-isolation}, is the only kind of non-serializable
behavior allowed. When referring to ``Repeatable Read,'' we will
follow these definitions, which, as we will soon show, are for a
property that can be offered in a highly available way.

@@

A stronger achievable property asks that a consistent cut spans both a
transaction's predicate reads (e.g., \texttt{SELECT WHERE}) as well as
its item reads.  We call this \textit{Predicate Cut
  Isolation}\footnote{Oracle provides an isolation level called
  ``Statement Level Read Consistency''~\cite{adya}; this is analogous
  to Predicate Cut Isolation at the level of a single operation within
  a multi-operation transaction. We do not consider this isolation
  model here as it is subsumed by standard Snapshot Isolation and we
  believe our discussion is easily extended to incorporate it.} and it
prohibits Phantoms. Predicate Cut Isolation is also achievable in HAT
systems via similar multi-versioning as above.

@@

Perplexingly, discussions of TA are absent from existing discussions
of weak isolation. This is perhaps again due to the single-node
context in which prior work was developed: on a single server (or a
fully replicated database), TA is achievable via lightweight locking
and/or local concurrency control over data items~\cite{gstore}. In
contrast, in a distributed environment, TA over arbitrary groups of
non-colocated items is considerably more difficult to achieve with
high availability: servers need to know when all of a transaction's
updates are present on their respective replicas. However, replicas do
not need to agree on when to reveal a new value to clients, which
would require consensus; accordingly, TA only requires reliable
broadcast (with some additional metadata for clients) and is
achievable in HAT systems. We omit a full discussion of the algorithm
and instead refer the reader to the extended version of this
paper~\cite{hat-tr}.  space{.5em}

@@

In this section, we discuss several subtleties in our results,
specifically addressing model composition, transactional atomicity
versus linearizability, and stickiness requirements.

@@

\vspace{.5em}\noindent\textbf{Linearizability and Transactional
  Atomicity} The relationship between linearizability and
transactional atomicity is non-obvious. Transactional atomicity
dictates that writes to multiple keys across multiple servers are made
visible to readers all at once, while linearizability dictates that
writes to a single key on multiple servers are made visible to all
readers at once---what is different? First, in linearizable (and safe
and regular) systems, writes are made visible to all clients
\textit{immediately} after they finish. With transactional atomicity,
there is no recency guarantee. Second, in linearizabile systems, all
clients see all writes at the same time. With transactional atomicity,
clients may see writes at different times depending on which replicas
they contact. We are not aware of an analogous model in the
distributed systems literature. Accordingly, despite apparent
similarities, transactional atomicity is incomparable with and much
cheaper (by availability standards) than linearizability.

### 2013-04-01 writeup/intro.tex
editing pass

The prior decade saw a large shift in the design of large-scale
database backends, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation,
  gray-virtues} to the widespread adoption of loosely-consistent
distributed key-value stores~\cite{bigtable, pnuts,
  dynamo}. Instrumental in this trend was the 2000 introduction of
Brewer's CAP Theorem, which stated that a highly available system
cannot provide strong consistency guarantees in the presence of
network partitions~\cite{brewer-slides}. However, as formally
proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability~\cite{herlihy-art}, the ability to read the most
recent write to a data item that is replicated across servers. Yet
this result---regarding a fairly narrow form of distributed
consistency---is often conflated with ACID database
properties~\cite{brewer-slides, hn, foundation-article}, leading to
substantial confusion about the relationships between replica
consistency, transactional isolation, and high availability. However,
transactions \textit{are} useful for programmers as evidenced by the
recent resurgence of transactional data stores such as Google's
Spanner~\cite{spanner}. Unfortunately, these systems are not highly
available in the presence of partitions~\cite{walter, mdcc,
  foundation-article, krikellas-bargain, eiger}. (TODO:ADD more)

Indeed, serializable transactions, the gold standard of traditional
ACID databases, is indeed not achievable given a requirement for high
availability in the presence of
partitions~\cite{davidson-survey}. However, database systems have a
long tradition of providing weaker isolation and consistency
guarantees~\cite{adya, ansicritique, gray-isolation,
  gray-virtues} (TODO: kemme). Due to their concurrency and overall performance
benefits, weak isolation models are widely employed, overwhelmingly as
the default setting in today's traditional ``ACID'' and up-and-coming
``NewSQL'' databases and often as the maximum guarantee offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics of current highly available systems. This raises a natural
question: which of these alternative semantics can be provided while
remaining highly available?

Unfortunately, the relationship of the CAP Theorem and ACID semantics
has not been well explored. We have a strong understanding of weak
isolation in the single-server context from which it
originated~\cite{adya, ansicritique, gray-isolation} and many papers
offer techniques for providing distributed
serializability~\cite{bernstein-book, hstore, kemme-classification,
  spanner, daudjee-session, krikellas-bargain, calvin} or
snapshot isolation~\cite{daudjee-snapshot,generalizedsnapshot,
  kemme-snapshot, walter}. The distributed computing, programming
language, and parallel hardware literature also contains many
consistency models for single operations on replicated
objects~\cite{pnuts, herlihy-art,  eiger, cac, sessionguarantees,
  ceri-mechanism, chen-mechanism}. However, the literature has largely
ignored semantic guarantees for multiple operations operating on
multiple data items in a highly available distributed environment.

Our main contributions are as follows. We fill gap between, on one
hand, the plethora of previously proposed isolation and consistency
models, and, on the other hand, the goal of high availability.  We
classify which, among the wide array of models, are highly available,
denoting them as {\em Highly Available Transactions (HATs)}. In doing
so, we demonstrate that many \textit{implementations} of HAT semantics
are not highly available but that their unavailability is not
\textit{inherently} required. Our investigation shows, somewhat
surprisingly, that besides serializability, only Snapshot Isolation
(SI) and Repeatable Reads (RR) are unavailable, with most other
isolation levels being available. We also show that many of the weaker
data consistency models from distributed systems are HAT-compliant.

>>>>

The prior decade saw a large shift in the design of large-scale
database backends, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
widespread adoption of loosely-consistent distributed key-value
stores~\cite{bigtable, pnuts, dynamo}. Instrumental in this trend was
the 2000 introduction of Brewer's CAP Theorem, which stated that a
highly available system cannot provide strong consistency guarantees
in the presence of network partitions~\cite{brewer-slides}. However,
as formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability~\cite{herlihy-art}, the ability to read the most
recent write to a data item that is replicated across servers. Yet
this result---regarding a fairly narrow form of distributed
consistency---is often conflated with ACID database
properties~\cite{brewer-slides, hn, foundation-article}, leading to
substantial confusion about the relationships between replica
consistency, transactional isolation, and high availability. However,
transactions \textit{are} useful for programmers as evidenced by the
recent resurgence of transactional data stores such as Google's
Spanner~\cite{spanner}. Unfortunately, these systems are not highly
available in the presence of partitions~\cite{ foundation-article,
  mdcc, krikellas-bargain, eiger, walter}. (TODO:ADD more)

Indeed, serializable transactions, the gold standard of traditional
ACID databases, is indeed not achievable given a requirement for high
availability in the presence of
partitions~\cite{davidson-survey}. However, database systems have a
long tradition of providing weaker isolation and consistency
guarantees~\cite{adya, ansicritique, gray-virtues, gray-isolation,
  kemme-thesis}. Due to their concurrency and overall performance
benefits, weak isolation models are widely employed, overwhelmingly as
the default setting in today's traditional ``ACID'' and up-and-coming
``NewSQL'' databases and often as the maximum guarantee offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics of current highly available systems. This raises a natural
question: which of these alternative semantics can be provided while
remaining highly available?

Unfortunately, the relationship of the CAP Theorem and ACID semantics
has not been well explored. We have a strong understanding of weak
isolation in the single-server context from which it
originated~\cite{adya, ansicritique, gray-isolation} and many papers
offer techniques for providing distributed
serializability~\cite{bernstein-book, spanner, daudjee-session,
  hstore, krikellas-bargain, calvin, kemme-classification} or snapshot
isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
  walter}. The distributed computing, programming language, and
parallel hardware literature also contains many consistency models for
single operations on replicated objects~\cite{ ceri-mechanism,
  chen-mechanism, pnuts, herlihy-art, eiger, cac, sessionguarantees}. However, the literature has largely ignored semantic guarantees for
multiple operations operating on multiple data items in a highly
available distributed environment.

Our main contributions are as follows. We fill the gap between, on one
hand, the plethora of previously proposed isolation and consistency
models, and, on the other hand, the goal of high availability.  We
classify which, among the wide array of models, are highly available,
denoting them as {\em Highly Available Transactions (HATs)}. In doing
so, we demonstrate that many \textit{implementations} of HAT semantics
are not highly available but that their unavailability is not
\textit{inherently} required. Our investigation shows, somewhat
surprisingly, that besides serializability, only Snapshot Isolation
(SI) and Repeatable Reads (RR) are unavailable, with most other
isolation levels being available. We also show that many of the weaker
data consistency models from distributed systems are HAT-compliant.

### 2013-04-01 writeup/model.tex
editing pass

To understand which guarantees can be provided with high availabilty,
we must first define what high availability means. In this section, we
will formulate a model that captures a range of availability models,
including ``true'' high availability, availablility with stickiness,
and several other useful variants.

Informally, highly available algorithms ensure ``always on'' operation
and guaranteed low latency. If users of a highly available system are
able to contact a (set of) server(s) in a system, they are guaranteed
a response; this means servers will not need to communicate with any
others. If servers are partitioned from one another, they do not need
to stall in order to provide a ``safe'' query response. This lack of
fast-path coordination also means that a highly available system also
provides low latency~\cite{abadi-pacelc}; in a wide-area setting, this
means that clients need not need to wait for cross-datacenter
communication. To properly describe whether a \textit{transactional}
system is highly available, we need to describe what servers a client
must contact as well as what kinds of responses a server can provide,
especially given the possibility of aborts.

>>>>

To understand which guarantees can be provided with high availabilty,
we must first define what high availability means. In this section, we
will formulate a model that captures a range of availability models,
including ``true'' high availability, availability with stickiness,
and several other useful variants.

Informally, highly available algorithms ensure ``always on'' operation
and guaranteed low latency. If users of a highly available system are
able to contact a (set of) server(s) in a system, they are guaranteed
a response; this means servers will not need to communicate with any
others. If servers are partitioned from one another, they do not need
to stall in order to provide a ``safe'' query response. This lack of
fast-path coordination also means that a highly available system also
provides low latency~\cite{abadi-pacelc}; in a wide-area setting, this
means that clients need not wait for cross-datacenter
communication. To properly describe whether a \textit{transactional}
system is highly available, we need to describe what servers a client
must contact as well as what kinds of responses a server can provide,
especially given the possibility of aborts.

### 2013-04-01 writeup/motivation.tex
eval
editing pass

Many of our discussions with practitioners---especially those
operating on public cloud infrastructure---as well as reports from
large-scale operators like Google~\cite{dean-keynote} confirm that
partition management is an important consideration service operators
today. As an example of an actual system rearchitecture resulting from
partitions, less than one year after its announcement, Yahoo!'s PNUTS
developers explicitly added support for weaker, highly available
operation. The engineers explained that ``strict adherence [to strong
  consistency] leads to difficult situations under network
partitioning or server failures...in many circumstances, applications
need a relaxed approach''~\cite{pnuts-update}.

@@

\subsection{Latency and Planet Earth}

@@

The previous section demonstrated that distributed systems must
address partitions and latency: what does this mean for distributed
databases? Database researchers and designers have long realized that
serializability is not achievable in a highly available
system~\cite{davidson-survey}, meaning it will be expensive in the
environments from Section~\ref{sec:motivation}. However, even a
single-node database, the coordination penalties associated with
serializability can be severe, manifesting themselves in the form of
decreased concurrency (and, subsequently, performance degradation,
scalability limitations, and, often, external aborts like
deadlocks)~\cite{gray-isolation}. Accordingly, o increase concurrency,
database systems offer a range of ACID properties weaker than
serializability.  Instead, databases offer a host of so-called
\textit{weak isolation} models that allow varying restrictions on the
space of schedules that are allowable by the system~\cite{adya,
  ansicritique, ansi-sql}. None of these weak isolation models
guarantees serializability, but, as we see below, the benefits of
these models are often considered to outweigh costs of possible
consistency anomalies that might arise from their use.

@@

It is unknown \textit{which} of these ACID (particularly isolation)
guarantees can be provided with high availability. The existing
algorithms for providing weak isolation are typically designed for a
single-node context and are, to the best of our knowledge, unavailable
due to a reliance on concurrency control mechanisms like locking that
are not resilient to partial failure
(Section~\ref{sec:evaluation}). Morever, we are not aware of any prior
literature that provides guidance as to the relationship between weak
isolation and high availability: prior work has examined the
relationship between serializability and high
availability~\cite{davidson-survey} and weak isolation on a single
server~\cite{adya, ansicritique, gray-isolation} but never weak
isolation and high availability taken together.  A primary goal in the
remainder of this paper is to understand which models are
HAT-compliant.

>>>>

Many of our discussions with practitioners---especially those
operating on public cloud infrastructure---as well as reports from
large-scale operators like Google~\cite{dean-keynote} confirm that
partition management is an important consideration for service operators
today. As an example of an actual system rearchitecture resulting from
partitions, less than one year after its announcement, Yahoo!'s PNUTS
developers explicitly added support for weaker, highly available
operation. The engineers explained that ``strict adherence [to strong
  consistency] leads to difficult situations under network
partitioning or server failures...in many circumstances, applications
need a relaxed approach''~\cite{pnuts-update}.

@@

\subsection{Latency and Planet Earth}
\label{sec:latency}

@@

The previous section demonstrated that distributed systems must
address partitions and latency: what does this mean for distributed
databases? Database researchers and designers have long realized that
serializability is not achievable in a highly available
system~\cite{davidson-survey}, meaning it will be expensive in the
environments from Section~\ref{sec:motivation}. However, even in a
single-node database, the coordination penalties associated with
serializability can be severe, manifesting themselves in the form of
decreased concurrency (and, subsequently, performance degradation,
scalability limitations, and, often, external aborts like
deadlocks)~\cite{gray-isolation}. Accordingly, to increase concurrency,
database systems offer a range of ACID properties weaker than
serializability.  Instead, databases offer a host of so-called
\textit{weak isolation} models that allow varying restrictions on the
space of schedules that are allowable by the system~\cite{adya,
  ansicritique, ansi-sql}. None of these weak isolation models
guarantees serializability, but, as we see below, the benefits of
these models are often considered to outweigh costs of possible
consistency anomalies that might arise from their use.

@@

It is unknown \textit{which} of these ACID (particularly isolation)
guarantees can be provided with high availability. The existing
algorithms for providing weak isolation are typically designed for a
single-node context and are, to the best of our knowledge, unavailable
due to a reliance on concurrency control mechanisms like locking that
are not resilient to partial failure
(Section~\ref{sec:evaluation}). Moreover, we are not aware of any prior
literature that provides guidance as to the relationship between weak
isolation and high availability: prior work has examined the
relationship between serializability and high
availability~\cite{davidson-survey} and weak isolation on a single
server~\cite{adya, ansicritique, gray-isolation} but never weak
isolation and high availability taken together.  A primary goal in the
remainder of this paper is to understand which models are
HAT-compliant.
